---
title: "Extract from Brenda"
#author: Sarah Torrence
date: December 20, 2020
output: github_document
---

Below is the code I used to extract and clean data from the brenda database in an easy to use format (data frame, data table).

```{r, message = FALSE}
library(brendaDb)
library(data.table)
library(tidyverse)
library(stringr)
library(splitstackshape)
library(janitor)
```

This function, `clean_column` is a helper function called several times in other functions below. It is used to clean up columns that have more than one value in each observation so that the data is in a tidy format, meaning each observation has one value in column.

```{r, warning = FALSE}
clean_column <- function(df, col_name){
  #cleans up a column that is a list of values so each cell has its own row
  
  #splitting up the values of the column
  df <- cSplit(df, splitCols = col_name , sep = ",", direction = "wide", drop = FALSE)
    
    #finding columns to input into the pivot_longer function
    param_columns <- names(df)
    columns <- param_columns[str_detect(param_columns, paste0("^",col_name,"_"))]
      
    #Transforming col_name into one column with one value for each row
    df <- df %>% pivot_longer(cols = columns, 
                                    names_to = "name", 
                                    values_to = paste0("new_",col_name), 
                                    values_drop_na = TRUE) %>% 
                               select(-name, -col_name)
}
```

The `extract_parameters` function is used to extract parameters from the brenda database. The function requires a specific EC number as an input as well as a list of the desired parameters from the brenda database. These parameters include km value, turnover number, ki value, pi value, ph optimum, ph range, temperature optimum, temperature range, specific activity and ic50. 

```{r}
extract_parameters <- function(ec, ls, parameters){
  #Extracts specified parameters for one EC, cleans up names and adds them to a new list
  
  #initializing the list
  param <- list()

  #iterate through all parameters needed
  for (i in parameters){
    
    #if the parameter is NA don't include it
    if (length(ls[[ec]][["parameters"]][[i]]) == 0 | is.na(ls[[ec]][["parameters"]][[i]])){
      break
    }
    else {
      #extract the parameter and clean up element names
      new_frame <- ls[[ec]][["parameters"]][[i]]
      names(new_frame)[2:4] <- c(i, "substrate", paste0(i,"_commentary"))
      
      #cleaning refID and proteinID so each row has one ID
      new_frame <- clean_column(new_frame,"refID")
      new_frame <- clean_column(new_frame,"proteinID")
      new_frame <- new_frame %>% select(refID = new_refID, proteinID = new_proteinID, everything())
      
      #append parameter to the list
      param <- append(param, list(new_frame))
    }
  }
  
  #return the list
  return(param)
}
```

`join_parameters` is used to take a list of extracted parameters (created using `extract_parameters` above) and join them into one data table or data frame that is easy to use for anaylsis. 

```{r}
join_parameters <- function(ls){
  #joins a list of parameter tables into one data table
  
  #converts each parameter into a data table
  ls <- lapply(ls, as.data.table)
  for (i in ls){
    setkey(i, proteinID, refID, substrate)
  }
  
  #join parameter data tables
  MergedDT <- Reduce(function(...) merge(..., all = TRUE, allow.cartesian = TRUE), ls)
  MergedDT <- clean_names(MergedDT)
}
```

Similarly to `extract_parameters`, `extract_organisms` extracts the organisms for an inputed EC number and cleans the data to get the list of organism information in a tidy data table or data frame format.

```{r}
extract_organisms <- function(ec, ls){
  #Extracts and cleans the organisms data for one EC
  
  #filtering for organisms
  organisms <- ls[[ec]][["organism"]][["organism"]]
  
  #cleaning variable names
  names(organisms)[2] <- "organism"
  
  #cleaning refID so each row has one ID
  organisms <- clean_column(organisms,"refID")
  organisms <- organisms %>% select(refID = new_refID, everything())
  
  organisms <- clean_names(organisms)
  organisms[, "protein_id"] <- lapply(organisms[, "protein_id"], as.integer)
  organisms <- as.data.table(organisms)
  
  return(organisms)
}
```

This function, `join_param_organisms` is pretty intuitive. It joins the data tables or data frames created by the above functions for parameters and organisms to get all the information for each EC number into one table.

```{r}
join_param_organisms <- function(parameters, organisms){
    #joins the parameters and organisms into one data table
  
    EC <- merge(organisms, parameters, by.x = c("protein_id","ref_id") , 
                 by.y = c("protein_id",  "ref_id"), all = TRUE, allow.cartesian = TRUE)
}
```

`one_EC` is a function that calls all the functions above to perform the process of extracting and cleaning parameter and organism information for one EC number inputted by the user. 

```{r}
one_EC <- function(ec, ls, parameters){
  #cleans, joins and outputs a data table for one EC number

  param <- extract_parameters(ec, ls, parameters)
  
  #if the ECs has no parameter information assign it to NA
  if (length(param) == 0){
    return(NA)
  }
  else {
    param <- join_parameters(param)
    
    org <- extract_organisms(ec, ls)
    
    EC <- join_param_organisms(param, org)
  }
}
```

The final function, `brenda_EC_data` takes the `one_EC` function above and performs it across all the desired EC numbers in the data set. It returns a tidy data table or data frame of all the information desired from the brenda database in an easy to use format for further analysis or processing.

```{r}
brenda_EC_data <- function(brenda_data, EC_list, parameters){
  #queries data from the brenda db for all desired ECs, processes the data and
  #outputs the data into one data table
  
  #querying data from the brenda Db
  ECs <- QueryBrenda(brenda_data, EC = EC_list)
  
  name <- list()
  out <- list()
  #iterating over all EC numbers and cleaning the data in a data table for each one
  for (i in 1:length(ECs)){
    o <- one_EC(names(ECs[i]),ECs, parameters)
    #if the EC is NA ignore it and move on
    if (is.na(o)){
      next
    }
    #appending each EC data table to a list
    name <- append(name, EC_list[i])
    out <- append(out, list(o))
  }
  names(out) <- name
  #merging all the data tables into one larger data table
  out_table <- rbindlist(out, use.names = TRUE, idcol = "ec", fill = TRUE)
  #filtering out rows where no valuable information is provided
  out_table <- out_table %>% filter((!is.na(km_value) & km_value != "additional information") |
                       (!is.na(turnover_number) & turnover_number != "additional information"))
  return(out_table)
}
```
