---
title: "How to Extract and Clean Data from Brenda"
#author: Sarah Torrence
date: November 19, 2020
output: github_document
---

If you have not yet installed the `brendaDb` package in R, run this code chuck to get the package installed on your machine.

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("brendaDb", dependencies=TRUE)
```

Here are the libraries I use.

```{r, message = FALSE}
library(brendaDb)
library(data.table)
library(tidyverse)
library(stringr)
library(splitstackshape)
library(janitor)
source("brenda_extract_functions.R")
```

You can use the `brendaDb` package to download the database and read it into R.

```{r, warning = FALSE, message = FALSE}
#downloading the database and reading it into R
brenda.filepath <- DownloadBrenda()
brenda <- ReadBrenda(brenda.filepath)
```

This process can take a while and must be done each time you want to use the database. Instead of having to constantly download the database each time you can write the data to a text file and read it in as a data frame or data table as shown below. 

```{r}
#writing the db to a text file
fwrite(brenda,file = "brenda.txt")

#reading in the db as a data.table
brenda_data <- fread("brenda.txt")
```

Here is an example query of how to extract meaningful data from the database. This example pulls out the information for the organism "Acetobacter pasteurianus" from the EC number 1.1.1.1. 

```{r, warning = FALSE}
example <- QueryBrenda(brenda_data, EC = "1.1.1.1", organisms = "Acetobacter pasteurianus")
ace <- example[["1.1.1.1"]][["parameters"]][["km.value"]]

ace %>% filter(fieldInfo == "acetaldehyde")
```
To be able to extract all the information desired from the brenda database, we can use the functions detailed [here](https://github.com/sktorre/brenda_research/blob/main/brenda_extract.md). The following is an example of using those functions to extract the parameters listed below for EC number 1.1.1.1.

```{r, warning = FALSE, message = FALSE}
parameters <- c("km.value", "turnover.number", "ki.value", "temperature.range", 
                "temperature.optimum", "ph.range", "ph.optimum")

brenda <- brenda_EC_data(brenda_data, "1.1.1.1", parameters)

head(brenda)
```

